CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems

### 2014-12-07 doc/journal/fpgapaperfinal/FPGAPaper/fpga2014paper.tex
testing my svn access
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4519 8e3573b8-cf2c-4f14-ef6d-137439e28b8b
changing paper template from ACM Conference to IEEETran. this also requires changing some of the Latex function calls. Also adding the section on placer optimization results.
git-svn-id: https://vtr-verilog-to-routing.googlecode.com/svn/branches/javeed@4521 8e3573b8-cf2c-4f14-ef6d-137439e28b8b

% THIS IS SIGPROC-SP.TEX - VERSION 3.1
% WORKS WITH V3.2SP OF ACM_PROC_ARTICLE-SP.CLS
% APRIL 2009
%
% It is an example file showing how to use the 'acm_proc_article-sp.cls' V3.2SP
% LaTeX2e document class file for Conference Proceedings submissions.
% ----------------------------------------------------------------------------------------------------------------
% This .tex file (and associated .cls V3.2SP) *DOES NOT* produce:
%       1) The Permission Statement
%       2) The Conference (location) Info information
%       3) The Copyright Line with ACM data
%       4) Page numbering
% ---------------------------------------------------------------------------------------------------------------
% It is an example which *does* use the .bib file (from which the .bbl file
% is produced).
% REMEMBER HOWEVER: After having produced the .bbl file,
% and prior to final submission,
% you need to 'insert'  your .bbl file into your source .tex file so as to provide
% ONE 'self-contained' source file.
%
% Questions regarding SIGS should be sent to
% Adrienne Griscti ---> griscti@acm.org
%
% Questions/suggestions regarding the guidelines, .tex and .cls files, etc. to
% Gerald Murray ---> murray@hq.acm.org
%
% For tracking purposes - this is V3.1SP - APRIL 2009

\documentclass{sig-alternate-2013}
\usepackage{graphicx}
\usepackage{subfigure}

\usepackage{epstopdf}
\usepackage{enumitem}
%\linespread{0.94}

\newfont{\mycrnotice}{ptmr8t at 7pt}
\newfont{\myconfname}{ptmri8t at 7pt}
\let\crnotice\mycrnotice%
\let\confname\myconfname%

\permission{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or commercial advantage and that copies bear this notice and the full citation on the first page. Copyrights for components of this work owned by others than the author(s) must be honored. Abstracting with credit is permitted. To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee. Request permissions from permissions@acm.org.}
\conferenceinfo{FPGA'14,}{February 26--28, 2014, Monterey, CA, USA. \\
{\mycrnotice{Copyright is held by the owner/author(s). Publication rights licensed to ACM.}}}
\copyrightetc{ACM \the\acmcopyr}
\crdata{978-1-4503-2671-1/14/02\ ...\$15.00.\\
http://dx.doi.org/10.1145/2554688.2554776}

\clubpenalty=10000
\widowpenalty = 10000

\begin{document}

% --- Author Metadata here ---
%\conferenceinfo{FPGA}{'14 Monterey, California USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

\title{CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems}
%
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.

\numberofauthors{2}
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Andre Hahn Pereira\\
	   \affaddr{Computer and Digital Systems Engineering Department - Escola Polit\'{e}cnica}\\
       \affaddr{University of S\~{a}o Paulo}\\
       \affaddr{S\~{a}o Paulo, SP}\\
       \affaddr{Brazil}\\
       \email{andre.hahn@usp.br}
% 2nd. author
\alignauthor
Vaughn Betz\\
       \affaddr{Department of Electrical and Computer Engineering}\\
       \affaddr{University of Toronto}\\
       \affaddr{Toronto, ON}\\
       \affaddr{Canada}\\
       \email{vaughn@eecg.utoronto.ca}
}
\date{18 December 2013}

\maketitle
\begin{abstract}
%
Interposer-based multi-FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. Such devices allow larger FPGA systems to be built than one monolithic die can accomodate and are now commercially available. An open question, however, is how efficient such systems are compared to a monolithic FPGA, as the number of signals passing between dice is reduced and the signal delay between dice is increased in an interposer system vs. a monolithic FPGA.

@@

\category{B.7.1}{Integrated Circuits}{Types and Design Style}[VLSI]
\category{B.7.2}{Integrated Circuits}{Design Aids}[Placement and Routing]
\terms{Algorithms, Design}
\keywords{FPGA; Silicon interposer; 2.5D ICs}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

%	- number of cuts
\subsection{Number of cuts}
\textit{Number of cuts} describes how many cuts are made to the interposer-based FPGA versus a monolithic die. If \textit{number of cuts} equals 1 then there are 2 dice, and so on. We investigate values of 1 and 3 (2 and 4 dice). The Virtex 7 family has members with \textit{number of cuts} $=$ 2 and 3. Figure \ref{fig:fpga} shows a sample architecture with one cutline.

@@

%   - % wires cut
\subsection{\% wires cut}
This variable describes and models the reduced connectivity between different dice by specifying the fraction of routing wires that are removed at the border between dice. For example, if a channel had 200 wires and \textit{\% wires cut} was 70, 140 of them would be cut and only 60 would pass through the interposer. Higher values of \textit{\% wires cut} make an interposer easier to manufacture, and can reduce the interposer delay by allowing all the microbumps linking dice to be placed near the die edge. However, the higher \textit{\% wires cut} is, the less routable the multi-die system becomes. As described in \ref{virtex7section}, the Virtex 7 family has \textit{\% wires cut} $=$ 77\%.

%   - Added delay
\subsection{Increased delay}
Interposer wires are longer and wider than on-die wires and have microbumps on each end. \textit{Increased delay} models the resulting larger delay when compared with wires which are internal to a die. A reasonable estimate for this variable is around $1ns$, as presented by Chaware et al\cite{xilinxTSV}.

%   - Interposer-Routing interface
\subsection{Interposer-Routing interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

\begin{enumerate}[label=(\alph*)]
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

@@

\subsection{Placer optimization}
\label{cad_enh_placer_subsection}
Placement is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. We modified the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

%   - Delay
\subsubsection{Placer timing cost}
The usual VPR timing cost is a (criticality-weighted) summation of the estimated delay (given the current placement) of every connection required by the circuit~\cite{timing2000}:

\begin{equation} \label{eq:timing_eq_full}
\begin{split}
Timing\_Cost = \sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times \\ 
criticality(i,j)
\end{split}
\end{equation}
where \textit{ij} denotes a connection from block i to block j that exists in the circuit netlist.

@@

\begin{equation} \label{eq:timing_cost}
\begin{split}
delay(i,j) =& delay(\Delta x_{ij}, \Delta y_{ij}) + \\
&times\_crossed(i,j) \times delay\_increase
\end{split}
\end{equation}

@@

%   - Wiring / cut cost
\subsubsection{Placer wiring cost}
The bounding box cost estimates the amount of wiring required for a net, based on the number of pins and size of the net's bounding box. VPR's original formulation is~\cite{betz1997vpr}:

\begin{equation} \label{eq:bbcost}
\begin{split}
wiring\_cost_{orig} = \sum_{n=1}^{N_{nets}} q(n) \times [\frac{bb_x(n)}{avg\_chanx\_W(n)} + \\
\frac{bb_y(n)}{avg\_chany\_W(n)}]
\end{split}
\end{equation}

@@

\begin{equation} \label{eq:total_wiring}
wiring\_cost = wiring\_cost_{orig} + cut\_cost
\end{equation}

@@

\begin{equation} \label{eq:cost2}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times bbHeight(i) \times times\_crossed(n)
\end{equation}

where 
\begin{equation} \label{eq:cprime}
\begin{split}
C' = \frac{ratio\_wires\_cut}{avg\_chany\_W(n)}
\end{split}
\end{equation}

@@

\subsection{Router optimization}
\label{sec:router_optimizations}
Once the Routing Resource Graph is modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. The router starts at a net source node and expands by discovering adjacent routing nodes until it reaches the destination node. At any point, the router knows the cost of the path taken to the current point and it uses a \textit{lookahead cost function} to estimate the resource and delay cost from the current node to the destination. Using this lookahead cost, the router can properly sort routing alternatives. The lookahead cost function in VPR is not aware of the interposer, and therefore it takes a long time to converge to a routing decision. To make the lookahead aware of the interposer, we modify the lookahead cost function by adding terms that account for extra \textit{interposer delay} and extra \textit{interposer nodes} that may exist on the route from the current node to the destination node.

The VPR router uses the following to calculate a cost for a routing node \textit{r} on the route from \textit{i} to \textit{j}:
\begin{equation} \label{eq:routing_cost}
\begin{split}
Cost(r) = Criticality(i,j) \times Delay_{Elmore}(r, topology) + \\ [1-Criticality(i,j)] \times Congestion(r)
\end{split}
\end{equation}

@@

\begin{equation} \label{eq:lookahead_cost}
\begin{split}
Total\_Cost(r) = \sum_{\forall l \in RT(i,r)} Cost(l) + \alpha \times ExpectedCost(r,j)
\end{split}
\end{equation}

@@

\begin{equation} \label{eq:expected_routing_cost}
\begin{split}
ExpectedCost(r,j) = Criticality_{fac} \times T_{delay} + \\(1 - Criticality_{fac}) \times ExpectedCongestionCost(r,j);
\end{split}
\end{equation}

$T_{delay}$ is an estimate of the delay that the router predicts it will see on the route from \textit{r} to \textit{j}. Since (\ref{eq:expected_routing_cost}) assumes a homogeneous FPGA, $T_{delay}$ does not include the additional delay some routes experience by going through the interposer. Thus, for interposer-based architectures, we replace $T_{delay}$ in (\ref{eq:expected_routing_cost}) with $T^\prime_{delay}$ to include the delay of expected interposer hops from \textit{r} to \textit{j}.
\begin{equation} \label{eq:tdelay}
\begin{split}
T^\prime_{delay} = T_{delay} + T_{interposer} \times ExpectedInterposerHops
\end{split}
\end{equation}

@@

%   - Effectiveness / results
\subsection{Effectiveness of the enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}) \cite{luu2014vtr}. The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

@@

\subsubsection{Effectiveness of placer optimizations}
\label{sec:effectiveness_of_placer_enhancements}
Enabling the enhancements described in Section~\ref{cad_enh_placer_subsection} improves the results .... haven't finished running these yet. So, <TODO EHSAN>: fill this section with actual results. For now, let's use a graph from Andre as a placeholder!
The graphs show that combining the timing with the bounding box optimizations led to an average of 20\% improvement on the area-delay product and the most significant contribution was made by incorporation of \textit{cut\_cost} into the wire cost function.

We believe the key to the good performance of the new placer cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalize bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{comparison_minW.eps}
\caption{Minimum channel width vs. \textit{\% wires cut}.}
\label{fig:comparison_minW}
\end{figure}

\subsubsection{Effectiveness of router optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} greatly reduce the runtime of the router. Figure~\ref{fig:lookahead_runtime} shows the geomean of the router runtime for the 8 largest VTR circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture results}
\label{resultsSection}

@@

%ACKNOWLEDGMENTS are optional
\section{Acknowledgments}

This work was supported by a Ci\^{e}ncia sem Fronteiras scholarship from CNPq - Brazil and the NSERC/Altera Industrial Research Chair in Programmable Silicon.
%
\bibliographystyle{abbrv}
\bibliography{fpga2014paper}

\balancecolumns
% That's all folks!
\end{document}

>>>>

%% bare_jrnl.tex
%% V1.4
%% 2012/12/27
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8 or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

\documentclass[journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[journal]{../sty/IEEEtran}

\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{epstopdf}
\usepackage{enumitem}
\usepackage{cite}
\usepackage[cmex10]{amsmath}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
% Do not put math or special symbols in the title.
\title{CAD and Routing Architecture for Interposer-Based Multi-FPGA Systems}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{
		% 1st author
		Ehsan~Nasiri,
		% 2nd author
		Javeed~Shaikh,
		% 3rd author
		Andre~Hahn~Pereira,
		% 4th author
        and~Vaughn~Betz,~\IEEEmembership{Member,~IEEE,}%
\thanks{E. Nasiri is with Altera Corporation, Toronto Technology Center, Toronto, Ontario, Canada (email: enasiri@altera.com)}% 
\thanks{J. Shaikh is with Google, Mountain View, California, USA (email: jvds@google.com)}%
\thanks{A. H. Pereira is with the Computer and Digital Systems Engineering Department - Escola Polit\'{e}cnica, University of S\~{a}o Paulo, S\~{a}o Paulo, SP, Brazil (email: andre.hahn@usp.br)}%
\thanks{V. Betz is with the Department of Electrical and Computer Engineering, University of Toronto, Toronto, Ontario, Canada (email: vaughn@eecg.utoronto.ca)}%
\thanks{Manuscript received January 19, 2014.}}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations in the abstract or keywords.
\begin{abstract}
Interposer-based multi-FPGA systems are composed of multiple FPGA dice connected through a silicon interposer. Such devices allow larger FPGA systems to be built than one monolithic die can accomodate and are now commercially available. An open question, however, is how efficient such systems are compared to a monolithic FPGA, as the number of signals passing between dice is reduced and the signal delay between dice is increased in an interposer system vs. a monolithic FPGA.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Algorithms, Design, FPGA, Silicon interposer, 2.5D ICs. 
\end{IEEEkeywords}

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture Models}
\label{archSection}
%   - Define parameters
To properly model a silicon interposer FPGA, we use version 7.0 of the popular FPGA exploration toolset, Verilog-to-Routing (VTR)~\cite{luu2014vtr}. The logic synthesis portions of the flow (ODIN II and  ABC) are left untouched, while we modify the placement and routing portion of the flow (VPR~\cite{betz1997vpr}) to model and optimize for interposer-based FPGAs. The modifications are made in such a way that they require no changes to any of the input files, so one can experiment with interposer-based FPGAs with any existing benchmark circuits and any existing VPR-format FPGA architecture description simply by specifying appropriate command-line parameters.

@@

%	- number of cuts
\subsection{Number of Cuts}
\textit{Number of cuts} describes how many cuts are made to the interposer-based FPGA versus a monolithic die. If \textit{number of cuts} equals 1 then there are 2 dice, and so on. We investigate values of 1 and 3 (2 and 4 dice). The Virtex 7 family has members with \textit{number of cuts} $=$ 2 and 3. Figure \ref{fig:fpga} shows a sample architecture with one cutline.

@@

%   - % wires cut
\subsection{\% Wires Cut}
This variable describes and models the reduced connectivity between different dice by specifying the fraction of routing wires that are removed at the border between dice. For example, if a channel had 200 wires and \textit{\% wires cut} was 70, 140 of them would be cut and only 60 would pass through the interposer. Higher values of \textit{\% wires cut} make an interposer easier to manufacture, and can reduce the interposer delay by allowing all the microbumps linking dice to be placed near the die edge. However, the higher \textit{\% wires cut} is, the less routable the multi-die system becomes. As described in \ref{virtex7section}, the Virtex 7 family has \textit{\% wires cut} $=$ 77\%.

%   - Added delay
\subsection{Increased Delay}
Interposer wires are longer and wider than on-die wires and have microbumps on each end. \textit{Increased delay} models the resulting larger delay when compared with wires which are internal to a die. A reasonable estimate for this variable is around $1ns$, as presented by Chaware et al\cite{xilinxTSV}.

%   - Interposer-Routing interface
\subsection{Interposer-Routing Interface}
\label{sec:interposer-routing-interface}
In a simple architecture, the vertical routing wires that can not connect to an interposer wire are dangling. Since interposer wires are scarce, we want to be able to study architectures in which the interposer wires are more accessible. To do so, we have enhanced the router to support the following interposer-routing interface options.

\begin{enumerate}
\item Fan-in Transfer for Interposer Wires\\
For an architecture with \textit{\% wires cut} $=0$, all vertical routing wires have a corresponding interposer wire that connects them to the adjacent die. Practical interposers, however, can only accommodate a fraction of all vertical routing wires. When \textit{fanin transfer} option is \textit{on}, we place a multiplexer at the input of each available interposer wire and connect some adjacent ``dangling" vertical routing wires to the multiplexer. This allows for additional flexibility in routing by ensuring that \textit{every} vertical routing wire could \textit{potentially} drive an interposer wire through a multiplexer. As an example, if \textit{\% wires cut} $=75\%$, 3 of every 4 interposer wires are cut. Therefore, we place a 4-to-1 multiplexer at the input of the one interposer wire that is not cut, and we connect the 3 vertical routing wires that do not have a corresponding interposer wire to the multiplexer.

@@

\subsection{Placer Optimization}
\label{cad_enh_placer_subsection}
Placement is crucial in mitigating the impact of the reduced wiring and increased delay when crossing between dice; a good placement should minimize the number of signals crossing between dice, particularly time-critical ones. VPR uses 2 different costs as the metrics for its placer algorithm: the timing cost and the bounding box (wiring) cost. We modified the placer's bounding box and timing costs to account for the increased latency to cross the interposer and for the reduced wire capacity close to the cutlines, as the wire availability becomes more sparse.

%   - Delay
\subsubsection{Placer Timing Cost}
The usual VPR timing cost is a (criticality-weighted) summation of the estimated delay (given the current placement) of every connection required by the circuit~\cite{timing2000}:

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_eq_full}
Timing\_Cost=\sum_{\forall i, j \subset circuit} delay(\Delta x_{ij},\Delta y_{ij}) \times \\ 
criticality(i,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

where \textit{ij} denotes a connection from block i to block j that exists in the circuit netlist.

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:timing_cost}
delay(i,j) = delay(\Delta x_{ij}, \Delta y_{ij}) + \\
times\_crossed(i,j) \times delay\_increase
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

%   - Wiring / cut cost
\subsubsection{Placer Wiring Cost}
The bounding box cost estimates the amount of wiring required for a net, based on the number of pins and size of the net's bounding box. VPR's original formulation is~\cite{betz1997vpr}:

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:bbcost}
wiring\_cost_{orig} = \sum_{n=1}^{N_{nets}} q(n) \times [\frac{bb_x(n)}{avg\_chanx\_W(n)} + \\
\frac{bb_y(n)}{avg\_chany\_W(n)}]
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{equation}
\label{eq:total_wiring}
wiring\_cost = wiring\_cost_{orig} + cut\_cost
\end{equation}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:cost2}
cut\_cost = \sum_{n=1}^{N_{nets}} C' \times bbHeight(i) \times times\_crossed(n)
\end{multline}

where 
\setlength{\arraycolsep}{0.0em}
\begin{equation}
\label{eq:cprime}
C' = \frac{ratio\_wires\_cut}{avg\_chany\_W(n)}
\end{equation}
\setlength{\arraycolsep}{5pt}

@@

This cost function performs better than several others as shown in \cite{hahn2014cad}, and we believe the key to the good performance of this cost functions is that \textit{cut\_cost} produces gradual gains as bounding boxes crossing the cutline shrink to be closer and closer to being captured entirely on one side of the cutline. Consider for example the 3 bounding boxes shown in Figure \ref{fig:bb_illustration}. Note that bounding box (a) and (b) both cross the cutline, but (b) is penalized less by (\ref{eq:cost2}) because bounding box (b) is mostly on the lower side of the cutline; it is more likely that later placement changes will result in the bounding box moving entirely below the cutline, reducing interposer wiring demand. Cost function (\ref{eq:cost2}) penalizes bounding box (a) more than (b) and (b) more than (c) to guide placement to gradually move bounding boxes to one side or the other of a cutline.

\subsection{Router Optimization}
\label{sec:router_optimizations}
Once the Routing Resource Graph is modified as detailed in Section \ref{archSection}, the VPR router adapted automatically to the interposer architecture. The router starts at a net source node and expands by discovering adjacent routing nodes until it reaches the destination node. At any point, the router knows the cost of the path taken to the current point and it uses a \textit{lookahead cost function} to estimate the resource and delay cost from the current node to the destination. Using this lookahead cost, the router can properly sort routing alternatives. The lookahead cost function in VPR is not aware of the interposer, and therefore it takes a long time to converge to a routing decision. To make the lookahead aware of the interposer, we modify the lookahead cost function by adding terms that account for extra \textit{interposer delay} and extra \textit{interposer nodes} that may exist on the route from the current node to the destination node.

The VPR router uses the following to calculate a cost for a routing node \textit{r} on the route from \textit{i} to \textit{j}:
\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:routing_cost}
Cost(r) = Criticality(i,j) \times Delay_{Elmore}(r, topology) \\ + [1-Criticality(i,j)] \times Congestion(r)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:lookahead_cost}
Total\_Cost(r) = \sum_{\forall l \in RT(i,r)} Cost(l) + \alpha \times ExpectedCost(r,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

\setlength{\arraycolsep}{0.0em}
\begin{multline}
\label{eq:expected_routing_cost}
ExpectedCost(r,j) = Criticality_{fac} \times T_{delay} + \\(1 - Criticality_{fac}) \times ExpectedCongestionCost(r,j)
\end{multline}
\setlength{\arraycolsep}{5pt}

$T_{delay}$ is an estimate of the delay that the router predicts it will see on the route from \textit{r} to \textit{j}. Since (\ref{eq:expected_routing_cost}) assumes a homogeneous FPGA, $T_{delay}$ does not include the additional delay some routes experience by going through the interposer. Thus, for interposer-based architectures, we replace $T_{delay}$ in (\ref{eq:expected_routing_cost}) with $T^\prime_{delay}$ to include the delay of expected interposer hops from \textit{r} to \textit{j}.

\begin{multline}
\label{eq:tdelay}
T^\prime_{delay} = T_{delay} + T_{interposer} \times ExpectedInterposerHops
\end{multline}
\setlength{\arraycolsep}{5pt}

@@

%   - Effectiveness / results
\subsection{Effectiveness of the Enhancements}
\label{sec:CADeffect}
For all experiments in this paper, the remaining architecture parameters of the FPGA are taken from the ``flagship" architecture of the VTR project (\textit{k6\_frac\_N10\_mem32K\_40nm.xml}) \cite{luu2014vtr}. The parameters of this architecture are in line with both current commercial FPGAs and academic research into best practices. It consists of logic clusters with 10 fracturable 6-LUTs per block (N=10, k=6), and also includes 32kb RAM blocks and DSP blocks configurable to perform 9x9, 18x18 or 36x36 multipliers. The delay values in the architecture are taken from 40 nm circuit simulations and 40 nm commercial FPGAs. It uses unidirectional routing, with all wire segments having length L = 4.

@@

\subsubsection{Effectiveness of Placer Optimizations}
\label{sec:effectiveness_of_placer_enhancements}
Enabling the enhancements described in Section~\ref{cad_enh_placer_subsection} improves the minimum routable channel width and circuit speed. As shown in Figure~\ref{fig:comparison_minW}, for architectures with low \textit{\% wires cut} (high interposer capacity), the placer optimizations are only marginally beneficial compared to the original VPR placer since the interposer wires are not scarce; but as more wires are cut and we reach the realistic interposer capacity of commercial FPGAs, the placer optimizations provide higher and higher reduction in minimum channel width. When 80\% of wires are cut, the placer optimizations result in 11.3\% reduction in minimum channel width.

\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{placer_opt_minChW.eps}
\caption{Impact of placer optimizations on minimum channel width.}
\label{fig:comparison_minW}
\end{figure}

The optimizations also help the placer to place related entities on the same die to reduce the number of inter-die signal crossings and hence increase circuit speed. At 80\% wires cut, the placer optimizations result in 8.4\% higher circuit speed.
\begin{figure}[!htbp]
\centering
\includegraphics[width=\linewidth]{placer_opt_fmax.eps}
\caption{Impact of placer optimizations on circuit speed.}
\label{fig:comparison_minW}
\end{figure}

\subsubsection{Effectiveness of Router Optimizations}
\label{sec:effectiveness_of_router_enhancements}
The router optimizations described in Section~\ref{sec:router_optimizations} greatly reduce the runtime of the router. Figure~\ref{fig:lookahead_runtime} shows the geomean of the router runtime for the 8 largest VTR circuits for different \textit{\% wires cut} values. On average, the router runs 31 times faster and finds identical minimum channel widths for all circuits.

@@

%---------------------------------------------------------------
%---------------------------------------------------------------
%---------------------------------------------------------------
\section{Architecture Results}
\label{resultsSection}

@@

% use section* for acknowledgement
\section*{Acknowledgment}
This work was supported by a Ci\^{e}ncia sem Fronteiras scholarship from CNPq - Brazil and the NSERC/Altera Industrial Research Chair in Programmable Silicon. Some of our computations were performed on the GPC supercomputer at the SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for Innovation under the auspices of Compute Canada; the Government of Ontario; Ontario Research Fund - Research Excellence; and the University of Toronto.

\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
\bibliography{IEEEabrv,./fpga2014paper}

% that's all folks
\end{document}
